---
title: "class08"
author: "Aiden Quisenberry"
format: pdf
toc: TRUE
---

## Background

In today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data. 

## Data Import

The data is in CSV format:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)

```

A wee peak at the data
```{r}
head(wisc.df, 3)
```

>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```
569

>Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```
212

>Q3. How many variables/features in the data are sufficed with _mean?

```{r}
colnames(wisc.df)
length(grep(pattern = "_mean", colnames(wisc.df)))
```

We need to remove the `diagnoses` column before we do any further analysis of this dataset - we don't want to pass this to PCA etc. We will save it as a seperate wee vector that we can use later to compare our findings to those of experts.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()`. We will use the optional arguemnet `scale=TRUE` here as the data columns/features/dimensions are on very different scales in the orginal dataset.

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
attributes(wisc.pr)
```

```{r}
library(ggplot2)
head(wisc.pr$x)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()

```


>Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

```{r}
summary(wisc.pr)
screeplot <-  ggplot
```
44.27% of original variance.

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Three PCs were required to capture 72.63% of the variance.

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Seven PCs were required to capture 90% of the orginal data variance.

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```
It appears that there is a wide amount of variance in the plot. When cross-checking it to known diagnosis' there appears to be a clear trend for positive variance in the diagnosed patients.

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

This plot shows a similar trend, where directional varience of PC3 is less attributive to diagnosis then the positive directional variance of PC1.

## Variance Explained

# Calculating the variance contributed by each component

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

# Variance explained by each principal component: pve
```{r}
pve <- pr.var / sum((wisc.pr$sdev)^2)
head(pve)
```

# Plot variance explained for each principal component

```{r}
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

# Alternative scree plot of the same data, note data driven y-axis

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

## ggplot based graph
```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

## Communicating PCA Results

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation[,1]
```

No, There do not seem to be any other features that seem to contribute more weight than concave.points_mean does to the PCA1.

## Heirarchical Clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
```

>Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust) +
abline(18, 0, col="red", lty=2)
```
At roughly 18

## Selecting number of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 5)
table(wisc.hclust.clusters, diagnosis)
```

>Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?

No, It seems like 4 is the best grouping of data. When you add another and split it into 5 however, you do get more groups of pureresponse. That is, 3 of the 5 groups return data falling into only one of the diagnosis's. 

## Using different methods

>Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
wisc.hclustsingle <- hclust(data.dist, method = "single")
plot(wisc.hclustsingle)
wisc.hclust <- hclust(data.dist, method = "complete")
plot(wisc.hclust)
wisc.hclustaverage <- hclust(data.dist, method = "average")
plot(wisc.hclustaverage)
wisc.hclustward <-  hclust(data.dist, method = "ward.D2")
plot(wisc.hclustward)
```
In this case,  `ward.D2` works better as it clumps the data into 2 groups. Since we are only testing to see whether 2 groups are different from each other, malignant or benign, it makes sense to use it to make those two clumps.

## Combining methods

The idea here is that I can take my new variables (i.e. the scores on the PCs `wisc.pr$x`) that are better descriptors of the data-set than the orginal features (i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering.

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```
```{r}
grps <- cutree(wisc.pr.hclust, k = 2)
table(grps)
```
I can now run `table()` with both my clustering `grps` and the expert `diagnosis`
>Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses? 

It does pretty well but alines about 10% of the diagnoses wrong. 

```{r}
table(grps)
table(diagnosis)
table(grps, diagnosis)
```

>Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.
## Prediction

It seems to do better as in it assignsthe groups with less false accuracy

We can use our PCA model for prediction of new un-seen cases: 
## Prediction
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

Our cluster "1" has 179 "M" diagnosis
Our cluster "2" has 333 "B" diagnosis

179 TP
24  FP
333 TN
33  FN

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity: TN/(TN + FP)

```{r}
333/(333+24)
```

>Q16. Which of these new patients should we prioritize for follow up based on your results?

All paitents within group 1 that have been diagnosed benign, as well as all paitents in group 2 that are diagnosed malignant. These two groupping could very well be false diagnoses which could prove harmful to the paitents.