---
title: "class12"
author: "Aiden Quisenberry PID:A19083392"
format: pdf
toc: TRUE
---

```{r}
mxl <- read.csv("MXL.csv")
head(mxl)
```
>Q5

```{r}
gg_count <- sum(mxl$Genotype..forward.strand. == "G|G")

gg_count

genotypes <- table(mxl$Genotype..forward.strand.)

proportions <- (genotypes/nrow(mxl))

proportions*100

```

## Section 2: Initial RNA-Seq analysis

```{r}
fast1 <- read.csv("HG00109_1.fastq")
fast2<- read.csv("HG00109_2.fastq")
head(fast1)
tail(fast1)
head(fast2)
tail(fast2)
```

>Q13

# determining how many sample size and expression levels for certain genotypes.
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```
Sample Size for each genotype
```{r}
exprt <- table(expr$geno)
exprt
```
Median Expression Levels for each Genotype
```{r}
library(dplyr)

medians <- expr %>%
  group_by(geno) %>%
  summarise(medexpr = (median(exp, na.rm = T)))
medians
```

Dataframe for genotypes and their corresponding median expression.
```{r}
genexp <- data.frame(exprt, medians$medexpr)
genexp
```

108 individuals had AA genotype and a median expression of 31.25
233 individuals had AG genotype and a median expression of 25.06
121 individuals had GG genotype and a median expression of 20.07
>Q14

After seeing the boxplot for these data, you can see that the expression levels go significantly down when comparing Homo AA & Het AG to homo GG. This significant decrease in gene expression could help to explain the cause for the asthma phenotype observed in this individual.

Boxplot for this data

```{r}
library(ggplot2)

exprbxplt <- ggplot(expr, aes(geno, exp, fill = geno)) +
  geom_boxplot(notch=T)
exprbxplt
```

```{r}
summary(expr)
```
