---
title: "Class 5: Data Vix w/ ggplot"
author: "Aiden (PDF: A19083392)"
format: pdf
toc: TRUE
---

## Background

There are lots of ways to make plots in R. These include so-called "base R" (like the `plot()`) and add on packages like **ggplot**.

Let's make the same plot with these two graphics systems. We can use the inbuilt `cars` dataset:

```{r}
head( cars )
```
With "base R" we can simply: 

```{r}
plot(cars)
```


Now let's try ggplot. First I need to install the package using `install.packages("ggplot2")`. 

> **N.B** We never run an `install.packages` in a code chunk, otherwise we will re-install needlesely every time we render our document. 

Every time we want to use an add-on package we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot needs at least 3 things: 

1. The **data** i.e. stuff to plot
2. The **aes** or aesthetics that map the data to the llot
3. The **geom_** or the geometry i.e. the plot type such as points, lines, histograms etc..

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "Speed (mph)", 
       y = "Distance (ft)", 
       title = "Stopping Distance of Old Cars") +
  theme_minimal()
```

## Gener Expression Plot

Read some data on the effects of GLP-1 inhibitor (drug) on gene expression values.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on the page.

```{r}
ggplot(data = genes) +
  aes(x = Condition1, y = Condition2) +
  geom_point(col = "blue", alpha = 0.2)
```

Let's color by `State` up, down or no change.

```{r}
table(genes$State)
```

```{r}
ggplot(data = genes) +
  aes(x = Condition1, y = Condition2, col = State) +
  geom_point(size = 1) +
  scale_color_manual(values =
                       c("red",
                         "gray",
                         "green")) +
  labs(x = "Activity Before Drug", y = "Activity After Drug", title = "Gene Regulation After Drug Application") +
  theme_minimal()
```

## Going further with gapmider

Here we explore the famou`gapmider` dataset with some custom plots.

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. How many rows does this dataset have?

```{r}
nrow(gapminder)
ncol(gapminder)
```

> How many different continents are in this dataset?

```{r}
table(gapminder$continent)
```

Version 1 plot GDP vs LifeExp

```{r}
ggplot(data = gapminder) +
         aes(x = gdpPercap, y = lifeExp, col = continent) +
         geom_point(size = 1) +
  labs(x = "gdp", y = "life expectency")
```
I want to see a plot for each continent - in ggplot lingo this is called "faceting"

```{r}
ggplot(data = gapminder) +
         aes(x = gdpPercap, y = lifeExp, col = continent) +
         geom_point(size = 1) +
  labs(x = "gdp", y = "life expectency") +
  facet_wrap(~continent)
```

## First look at the dplyr package

Another add-on package with a function called `filter` that we want to use.

```{r}
library(dplyr)
 filtered <- filter(gapminder, year == 2007 | year == 1977)
ggplot(data = filtered) +
         aes(x = gdpPercap, y = lifeExp, col = continent) +
         geom_point() +
  facet_wrap(~year)
```

