---
title: "Class 18: Pertussis Mini Project"
author: "Aiden Quisneberry (PID:A19083392)"
format: pdf
toc: TRUE
---

## Background
Pertussis (a.k.a, Whooping Cough) is a common lung infection caused by the bacteria Bordatella Pertussis (a.k.a, B. Pertussis)

This disease mostly effects only humans and can infect all ages, the symptoms and long-term effects however are most severe for ages under 1 year and can lead to chronic illnes or even death.

The CDC track the number of [reported cases in US](https://github.com/MilesMcBain/datapasta):
We can "scrape" this data with the **datapasta** package.

```{r}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025
  ), cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
  )
)
```
>Q. Make a Plot of year vs cases

```{r}
library(ggplot2)
ggplot(cdc, aes(x = year, y = cases)) +
  geom_point() +
  geom_line() +
  labs( title = "Pertussis Cases by Year", x = "Year", y = "No. of Cases") +
  theme_minimal() +
  scale_y_continuous(labels = scales::label_comma())
```
>Q. Add some major milestones including the first wP vaccine roll-out (1946), the switch to the newer aP vaccine (1996), the COVID years (2020) 

```{r}
library(ggplot2)
ggplot(cdc, aes(x = year, y = cases)) +
  geom_point() +
  geom_line() +
  labs( title = "Pertussis Cases by Year in the US", x = "Year", y = "No. of Cases") +
  theme_minimal() +
  scale_y_continuous(labels = scales::label_comma()) +
  geom_vline(xintercept = c(1946, 1996, 2020), col=c("blue","red","yellow"), lty = 2 )
```

There were high case numbers in the pre 1940s, then in 1946 after the first wP vaccine was introduced the number of cases began to drastically reduce.

In 1996 the wP vaccine which required 1 shot was switched to an aP vaccine which required a booster every 10 years. Approximately 10 years later, there was a massive spike in cases. 

In 2020, the worldwide COVID pandemic caused a large drop in pertussis cases. This is likely caused by the social distancing and health regulations imposed to address the worldwide pandemic. During this time vaccines got a lot of controversy publicly.

We understand that aP induced protection wanes faster than the wP induced protection. Pulling information from other countries that switched vaccines, they also saw a correlated rise in cases approximately 10 years after its switch.

**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

# CMI-PB Project
[Computation Models of Immunity - Pertussis Boost project](https://www.cmi-pb.org/) aims to procide the scientific community with this very information.

They make their data available via JSON format APIs. We can read this is R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)
```

>Q. How many wP and aP individuals are in this `subject` table

```{r}
sum(subject$infancy_vac=="wP")
sum(subject$infancy_vac=="aP")
table(subject$infancy_vac)
```

>Q. What is the biological sex breakdown?

```{r}
table(subject$biological_sex)
```
>Q. In terms of race and gender, is this dataset representative of the US population?

```{r}
table(subject$race, subject$biological_sex)
```

Let's read some more database tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```
```{r}
head(specimen)
head(ab_titer)
```

To analyze this data we need to first "join" the different tables so we have all the data in one place and not spread across multiple tables.

We can use the `*_join` family of functions from **dplyr** to do this

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```


```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q. What Antibody isotypes are measured for these patrients?

```{r}
table(abdata$isotype)
```


> Q. What antigens are reported?

```{r}
table(abdata$antigen)
```

Let's focus on IgG isotype and make a plot of MFI_normalized for all antigens.

```{r}
igg  <- abdata |>
  filter(isotype == "IgG")
head(igg)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

> Q. Is there a difference for aP vs wP individuals with these values? 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

> Q. Is there a temporal response - i.e do values increase or decrease over time?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen,col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

## Focus on "PT" Pertussis Toxin antigen

```{r}
pt.igg.21 <-  igg |> filter(antigen == "PT",
                        dataset =="2021_dataset")
```


```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost,
      MFI_normalised,
      col=infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 14, lty = 2)
```

```{r, warning=FALSE}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost,
      MFI_normalised,
      col=infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line(alpha = .3) +
  geom_smooth(aes(group = infancy_vac),
              se = F,
              span = .4) +
  geom_vline(xintercept = 14, lty = 2)
```

