---
title: "Class06: R Functions"
author: "Aiden (PID:19083392)"
format: pdf
toc: TRUE
---

## Background

Functions are at the heart of using R. Everything we do involves calling and using functions (from data input, analysis to results output)

All functions in R have at least 3 things: 

1. A **name** the thing we use to call the function.
2. One or more input **arguements** that are `,` seperated.
3. The **body**, lines of code between curly brackets `{ }` that does the work of the function.

## A first function

Let's write a simple function that can add some numbers:

```{r}
add <- function(x) { 
  x + 1
  }
```

Let's try it out
```{r}
add(100)
```

```{r}
add(c(100, 200, 300))
```

Modify to be more useful and add more than just 1

```{r}
add <- function(x, y) {
  x + y
}
```

add(100,10)

>**N.B.** Input arguements can be either **required** or **optional**. The latter have a fall-back default that is specified in the function code with an equals sign.

```{r}
#add(x = 100, y = 200, z = 300)
```

## A Second Function

All functions in R look like this:

```
name <- function(arguement) {
 body
}
```

The `sample()` function in R takes a random sample of a vector or other datasat. It uses arguements that define the data, the number of rows you want to sample, whether you want to replace the sampled data points, and the probability of each vector having been sampled.

```{r}
sample(1:10, 4,)

```

>Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10, 12, replace = T)
```

>Q. Write the code to generate a random 12 nucleotide long DNA sequence?

```{r}
bases <- c("A","T","C","G") 
sample(bases, 12, replace = T)

```

>Q. Write a first version function called `generate_dna()` that generage a user specified length `n` random DNA sequence.

```{r}
generate_dna <- function(n=6) {
  bases <- c("A","T","C","G")
  sample(bases, n, T)
}
```

>Q. Modify your function to retunr a FASTA like sequence, so rather than [1] "G" "G" "C" "C" "C" "G" we want "GCAAT"

```{r}
generate_dna <- function(n=6) {
  bases <- c("A","T","C","G")
  seq <- sample(bases, n, T)
  paste(seq, collapse = "")
}
```

generate_dna(10)

>Q. Give the user an option to return FASTA format output sequence or standard multi-element vector format?

```{r}
generate_dna <- function(n=6, fasta = T) {
  bases <- c("A","T","C","G")
  dnaseq <- sample(bases, n, T)
  paste(dnaseq, collapse = "")
  
  if(fasta) {
  format <- paste(dnaseq, collapse = "")
  cat("hello...")
  }
  else(fasta)
  format <- paste(dnaseq, collapse = "")
  cat("Is it me you're looking for")
  return(dnaseq)
}
```

```{r}
generate_dna(10)
generate_dna(10, fasta = F)
```
## A new cool function
>Q. Write a function called `generate_protein` that generates a user specified length protein sequence in FASTA like format

```{r}
generate_protein <- function(n=6) {
  aa <- c("A","C","D","E","F","G","H",
          "I","K","L","M","N","P","Q",
          "R","S","T","V","W","Y")
  seq <- sample(aa, n, T)
  paste(seq, collapse = "")
}
```

generate_protein(10)

>Q. Use your new `generate_protein` function to generate all sequences between length 6 and 12 amino-acids in length and check if any of these are unique in nature (i.e. found in the NR database at NCBI)

We will use a `for()` loop:

```{r}
for(i in 6:12) {
  ans <- cat(">", i, sep="", "\n")
  cat(generate_protein(i), "\n")
}
```

