---
title: "Class07: Machine Learning 1"
author: "Aiden Quisenberry (A19083392)"
format: pdf
---

## Background
Today we will begin our exploration of some important machine learning methods, namely **clustering** and **dimensionality reduction**.

Let's make up some input data for clustering where we know what the natural "clusters" are.

the function `rnorm()` can be useful here.
```{r}
hist(rnorm(30, mean = 3, sd = 1))

```

>Q. Generate 30 random numbers centered at +3

```{r}
tmp <- c(rnorm(30, mean = 3),
         rnorm(30, mean = -3))

x <- cbind(x = tmp, y = rev(tmp))

plot(x)
```


## K-means clustering

K is the number clusters of sub-data you want from the data. The main function in "base R" for K-means clustering is called `kmeans()`

```{r}
km <- kmeans(x, centers = 2)
km
```

>Q. What component of the results object details the cluster sizes

```{r}
km$size
```

>Q. What component of the results object details the cluster centers?

```{r}
km$centers
```

>Q. What component of the results object details the cluster membership vector (i.e. our main result of which points lie in which cluster)?

```{r}
km$cluster
```
>Q. Plot our clustering results with points colored by cluster and also add the cluster centers as new points colored blue?

```{r}
plot(x, col=km$cluster)
points(km$centers, col = "blue", pch = 15)
```

>Q. Run `kmeans()` again and this time produce 4 clusters (call your result object `k4`) and make a results figure like above?

```{r}
k4 <- kmeans(x, centers = 4)

plot(x, col=k4$cluster)

points(k4$centers, col = "blue", pch = 15)

```

The metric
```{r}
km$tot.withinss
k4$tot.withinss
```

> Q. Let's try different number of K (center) from 1 to 30 and see what the best result is?


```{r}

ans <- NULL
for( i in 1:30 ) {
  ans <- c(ans, kmeans(x, centers = i)$tot.withinss)
}
```

```{r}
plot(ans, type = "o")
```
The right number of clusters is where the biggest drop in variation is between clusters, known as the `Scree Point`.. The goal is to minimize number of clusters while also minimizing the amount of variation in each cluster. In this case 2 clusters has the biggest drop in variation and is the right answer. 
**Key-point** K-means will impose culturing on data even if the cluster isnt there, if you ask it to.

## Hierarchical Clustering

The main function for Hierarchical Clustering is called `hclust()`
Unlike `kmeans()` (which does all the work for you), you can't just pass `hclust()` our raw input data. It needs a "distance matrix" like the one returned from the `dist()` function. 

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```


To extract our cluster membership vector from a `hclust()` result object we have to "cut" our tree at a given height to yield separate "groups"/"branches". 
```{r}
plot(hc)
abline(h = 8, col = "red", lty = 2)
```



To do this we use the `cutree()` function on our `hclust()` object:

```{r}
grps <- cutree(hc, h=8)
grps
```
```{r}
table(grps, km$cluster)
```

## PCA of UK food data

Import the dataset of food consumption in the UK:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

>Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

One solution to set the row names is to do it by hand ...

```{r}
rownames(x) <- x[,1]

```

To remove the first column I can use the munus `-` index trick

```{r}
x <- x[,-1]
```

A better way to do this is to set the row names to the first column with `read.csv`

```{r}
x <- read.csv(url, row.names = 1)
x
```

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances? 

I prefer argueing with the `read.csv` function by setting `row.names = 1`. This does not cause any issues when compiling or re-running code like the manual column removal does.

### Spotting major differences and trends

Is difficult even in this wee 17D datase

```{r}
barplot(as.matrix(x), beside = T, col = rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside = F, col = rainbow(nrow(x)))
```
>Q3: Changing what optional argument in the above barplot() function results in the following plot?

Changing "beside" from T to F

### Pairs plot and heatmaps

```{r}
pairs(x, col = rainbow(nrow(x)), pch = 16)
```
>Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

The pairs function plots each data point and compares it to the other points of the dataset. If a point lies on the diagonal for the plot it means they are reporting the same value. 

```{r}
library(pheatmap)

pheatmap(as.matrix(x) )
```
>Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Most of the foods between each country shows similar patterns, differences mainly lie in potatoes and other_meat

## PCA to the rescue

The main PCA function in "base R" is called `prcomp()`. This function wants the transpose of our food data as input (i.e. the foods as columns and the countries as rows)

```{r}
tx <- t(x)
pca <- prcomp(t(x))
```

```{r}
summary(pca)
```

```{r}
attributes(pca)
```

To make one of our main PCA result figures, we turn to `pca$x` the scores along our new PCs. This is called "PC plot" or "score plot" or "Ordination plot" ...

```{r}
my_cols <- c("orange","red","blue","darkgreen")
```


```{r}
library(ggplot2)
pca$x
ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point(col = my_cols)
```

The second major result figure is called a "loadings plot" or "variable contributions plot" or "weight plot"

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = rownames(pca$rotation)) +
  geom_col()
```

